Memories of migrations past: Sociality and cognition in dynamic, seasonal environments

### 2021-07-15 text/SocialMemoryMigration_FEE.tex
bibibbibib

%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
\section{}
For full guidelines regarding your manuscript please refer to \href{http://www.frontiersin.org/about/AuthorGuidelines}{Author Guidelines}.

As a primary goal, the abstract should render the general significance and conceptual advance of the work clearly accessible to a broad readership. References should not be cited in the abstract. Leave the Abstract empty if your article does not require one, please see \href{http://www.frontiersin.org/about/AuthorGuidelines#SummaryTable}{Summary Table} for details according to article type.

@@

Proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}. Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015}. These migrations can be considered ``tactical'', as they can occur - as an extreme simplification - purely as response to local conditions. Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration \citep{Abrahms2019}. This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether those are food, suitable habitat for breeding, or refuge from predators, that are highly seasonal and localized. This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017}.

Migratory species are often considered to be more vulnerable to environmental change, as a disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}. On the other hand, it has been argued that migratory species might be more resilient to disruptions due to their wide-ranging mobility \citep{Robinson2009}. Clearly, the ability to perform a migration without local cues is only possible if the behavior is hard-programmed or remembered. On the other hand, a strictly programmed behavior can be maladaptive if conditions change. Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.

@@

Here, we develop a social diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions. In formulating the model, our goal was to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior. This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience or fragility of a migratory population against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions, mirroring the effects of climate change.

@@

The drifting resource differs from the island resource in that the total amount of resource at any given time point is 1 ($\int_X h(x,t) dx = 1$). This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see appendix for details). Both types of resources are illustrated in figure \ref{fig_ResourceExamples}.

@@

We also found that the ability to maintain, learn and adapt a migration depends strongly on properties of the resource dynamics. In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space. Interestingly, in most stable patterns, the eventual targeted migration arrival time coincided with the \emph{peak}, rather than the beginning, of the resource dynamic. This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden (explosive). This is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths leading to a rapid and intense local green-up, or, in a more extreme case, when resources are linked to the short-duration early blooming phenology of very particular plants (\textbf{refs}).

@@

Finally, the migration timing and location parameter can be similarly straightforwardly estimated from movement data \citep{Cagnacci2015, Gurarie2019}. Thus, for example, \citet{Gurarie2019} explicitly estimated the ranging area, timing, and seasonal range locations for migratory caribou (\emph{Rangifer tarandus}), identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here as well as some trends in the timing of migration. Finally, we note that our underlying assumption of the migration ``urge'' is consistent with the strong endogenous programs to migrate that are well-documented. Birds in particular exhibit a seasonal restlessness known as \emph{Zugunruhe} \citep{Berthold1999, Helm2006}.

@@

The reference memory parameter $\kappa$ is, of course, impossible to observe directly. Our model does, however, allow us to explore in an heuristic way the conditions under which a strong cultural tendency to migrate with certain fixed patterns can help a population hedge against stochasticity \citep{Abrahms2019,Fagan2019b}.

@@

>>>>

%%% Leave the Abstract empty if your article does not require one, please see the Summary Table for full details.
\section{}
Seasonal migrations are a widespread and broadly successful strategy for animals to exploit

@@

Proximate causes, drivers and mechanism can vary considerably across and even within species \citep{Berthold1999, Shaw2016}. Some migrants follow a ``green wave'' of spring vegetation as it flowers across altitudinal or latitudinal gradients \citep{Bischof2012, Kolzsch2015, Merkle2016}. These migrations can be considered ``tactical'', as they can occur - as an extreme simplification - purely as response to local conditions. Other migrants perform long-distance migrations in anticipation that critical resources will be available at the time of arrival at the end point of migration \citep{Abrahms2019}. This second behavior involves the greatest trade-off between the costs of migration against the benefits of accessing resources, whether those are food, suitable habitat for breeding, or refuge from predators, that are highly seasonal and localized. This approach can be considered ``strategic'' in the sense that it is driven not by immediate cues but by an anticipation based on prior experience \citep{Bracis2017}.

Migratory species are often considered to be more vulnerable to environmental change, as a disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts \citep{Wilcove2008, Kauffman2021}. On the other hand, migratory species might be more resilient to disruptions due to their wide-ranging mobility \citep{Robinson2009}. Clearly, the ability to perform a migration without local cues is only possible if the behavior is hard-programmed or remembered. On the other hand, a strictly programmed behavior can be maladaptive if conditions change. Because the scenarios underlying migration are manifold and complex, mathematical modeling may provide some insights and help clarify where, when, and under what conditions we might migration behavior to emerge, to be adaptive, to be maladaptive, or to collapse.

@@

Here, we develop a social diffusion-advection model with memory to explore the resilience of a migratory population under various dynamic, seasonal resource distributions. In formulating the model, our goal was to identify the minimum set of movement and memory parameters required to generate an adaptive, migratory behavior. This includes the ability to learn to migrate from non-migratory initial conditions, simulating the release of naive animals in a seasonal environment \citep{Jesmer2018}, to lose the propensity to migrate if the resource distribution does not require it, also a commonly observed phenomenon \citep{Wilcove2008}, and to assess the resilience or fragility of a migratory population against changing resource distribution dynamics, including both stochasticity and trends in spatial and temporal distributions, mirroring the effects of climate change \citep{Park2020}.

@@

The drifting resource differs from the island resource in that the total amount of resource at any given time $\int_X h(x,t) dx = 1$. This property is attained by distributing the resource as a re-scaled beta distribution, where the shape and scale parameters vary sinusoidally in such a way as to make the standard deviations and means match the desired values of $m_x, m_t, s_x, s_t$ (see Supplementary Materials for details). Both types of resources are illustrated in figure \ref{fig_ResourceExamples}.

@@

We also found that the ability to maintain, learn and adapt a migration depends strongly on properties of the resource dynamics. In particular, the reinforcement of memory and foraging is strongest when patches are concentrated in time, but relatively large in space. Interestingly, in most stable patterns, the eventual targeted migration arrival time coincided with the \emph{peak}, rather than the beginning, of the resource dynamic. This indicates that the long-distance social migration behavior may be particularly reinforced when the targeted resource is very sudden. This is the case for the rapid green-up that occurs in high latitudes as snow recedes in tandem with extended day lengths leading to an intense green-up period \citep{Park2020} or, for example, when resources are linked to the short-duration early blooming phenology of very particular plants \citep{Post2007}.

@@

Migration timing and location parameter can be similarly straightforwardly estimated from movement data \citep{Cagnacci2015, Gurarie2019}. Thus, for example, \citet{Gurarie2019} explicitly estimated the ranging area, timing, and seasonal range locations for migratory caribou (\emph{Rangifer tarandus}), identifying the kind of inter-annual variation that is reflected in the stochastic scenarios explored here as well as some trends in the timing of migration. We note that our underlying assumption of the migration ``urge'' is consistent with the strong endogenous programs to migrate that are well-documented. Birds in particular exhibit a seasonal restlessness known as \emph{Zugunruhe} \citep{Berthold1999, Helm2006}.

@@

The reference memory parameter $\kappa$ is, of course, impossible to observe directly. Our model does, however, allow us to explore in an heuristic way the conditions under which a strong cultural tendency to migrate with certain fixed patterns can help a population hedge against stochasticity \citep{Abrahms2019,Fagan2019b}.  An extremely conservative behavior may be the best way to hedge against stochasticity (high $\kappa$ values in figure \ref{fig_stochasticity}) as it serves no benefit to change behavior based on recent experiences if they do not inform what might happen in the future. However, this extreme conservatism is, by definition, incapable of adapting when there is a consistent shift in resource disrtibution (figure \ref{fig_trendstochasticity}).  In cases where both processes are occurring, we did, as predicted, see a slight improvement in adaptability when some long-term reference memory was used to stability the inclination to adapt, visible as peaks in figure \ref{fig_trendstochasticity} at $\kappa$ values around 0.8.

@@

\clearpage
\appendix

\section{Supplementary material}
\normalsize

\subsection{Drifting resource}
The drifting resource function has the following properties

\begin{enumerate}
	\item The total amount of resource across space is constant throughout the year.
	\item At the beginning, middle, and end of the year the resource is uniformly distributed. 
	\item At some peak time $\mu_t < \tau/2$, the resource concentrates at a location $\mu_x < \chi$ with a spatial deviation $\sigma_x$ and a temporal deviation $\sigma_t$ (where $\tau$ is the length of the year and $\chi$ is the extent of the spatial domain).
	\item The resource peaks exactly symmetrically at time $\tau - \mu_t$ and location $-\mu_x$ with the same variances.
\end{enumerate}

To generate a resource with these properties, we distributing the resource in space as a beta distribution, where the two shape and scale parameters vary sinusoidally in such a way as to fulfill the criteria above.  Thus:

$$h(x,t, \theta) = \chi B(x/\chi, a(t, \theta), b(t, \theta))$$

where $\chi$ is the maximum value (domain) of $x$, $B(x, a, b)$ is the beta distribution, $\theta$ represents the set of parameters $t_r, x_r, \sigma_t, \sigma_x$, and the two shape parameters are given by:

$$a(t) = \frac{m}{s^2}( s^2 + m - m^2)$$
$$b(t, x', \sigma') = (m-1)\left(1 + \frac{m}{s}(m-1)\right) $$

where $m(t)$ and $s(t)$ describe the dynamic mean and variance of the resource peak.  These equations are solutions to the mean and variance of the beta distribution, $\mu = \alpha/(\alpha + \beta)$, $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$.

The means and variances themselves are Gaussian pulses, with the mean peaking at $\mu_x$ at time $\mu_t$ with standard deviation $\sigma_t$ and at $-\mu_x$ at time $\tau - \mu_t$ and the standard deviation pulsing from $2 \chi/\sqrt{12}$ (corresponding to a uniform distribution over the domain $-\chi$ to $\chi$) at times 0, $\tau/2$ and $\tau$ down to $\sigma_x$ at $t_r$ and $\tau - t_r$, with standard deviation (in time) $\sigma_t$.
